PyPy's Approach to Virtual Machine Construction

### 2016-08-18 talk/dls2006/paper.tex
remove a sentence -- enough to get the document to fit onto 12 pages.
\label was at the wrong place
Fix table labels.
Add another point

\begin{figure}
\begin{verbatim}
def ll_dict_insertclean(d, key, value, hash):
  entry = ll_dict_lookup_clean(d, hash)
  ENTRY = lltype.typeOf(entry).TO
  entry.value = value
  entry.key = key
  if hasattr(ENTRY, 'f_hash'): 
    entry.f_hash = hash
  if hasattr(ENTRY, 'f_valid'):    
    entry.f_valid = True
  if hasattr(ENTRY, 'f_everused'): 
    entry.f_everused = True
  d.num_items += 1
  d.num_pristine_entries -= 1
\end{verbatim}
\label{fig_insertclean}
\caption{built-in dict insertion helper.}
\end{figure}

@@

%%% XXX the first sentence seems to come a bit out of the blue.

We found in-process debugging sufficient and did not implement dumping
of any intermediate step to disk.  Figure \ref{fig_screenshot}
shows the control flow graph
obtained for a simple function -- this is a screenshot from our graph
viewer, used for debugging; basic block placement is performed by
Graphviz \cite{graphviz}.

@@

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.

Its lattice order (actually a join-semilattice only) is extremely
simple, because most actual analysis is delayed to the next phase, the
type inference engine.  The objects are either \textit{Variables},
which are pure placeholders for entirely unknown values, or
\textit{Constants} with a concrete Python object as value.  The order
places Variable as the top, and keeps all \textit{Constants}
unordered.  Thus if two different constants merge during abstract
interpretation, we immediately widen them to Variable.

@@

These types are ordered to form a lattice -- we generally use its
structure of join-semilattice only, but we will see in section
\ref{precision} a use case for the \textit{meet}. Pbcs form a
classical finite set-of-subsets lattice, class annotations are ordered
by inheritance.  We have left out a number of other annotations that
are irrelevant for the basic description of the annotator and
straightforward to handle: $Dictionary$, $Tuple$, $Float$,
$UnicodePoint$, $Iterator$, etc.  The complete list and order is
described in \cite{T}.

@@

\begin{table*}
% the layout of this table isn't wonderful, but it's probably OK.
\centering
\label{perfsumm}
\begin{tabular}{|l|c|c|} \hline
\textbf{Interpreter} & 
\textbf{Richards, Time/iteration} &
\textbf{Pystone, Iterations/second} \\ \hline
CPython 2.4.3                &   789ms  (1.0x) & 40322  (1.0x) \\ \hline
pypy-c                       &  4269ms  (5.4x) &  7587  (5.3x) \\ \hline
pypy-c-thread                &  4552ms  (5.8x) &  7122  (5.7x) \\ \hline
pypy-c-stackless             &  5121ms  (6.5x) &  6060  (6.7x) \\ \hline
pypy-c-gcframework           &  6327ms  (8.0x) &  4960  (8.1x) \\ \hline
pypy-c-stackless-gcframework &  8743ms (11.0x) &  3755 (10.7x) \\ \hline
pypy-llvm-c                  &  3797ms  (4.8x) &  7763  (5.2x) \\ \hline
pypy-llvm-c-prof             &  2772ms  (3.5x) & 10245  (3.9x) \\ \hline
\end{tabular}
\caption{summary of interpreter performance.}
\end{table*}

@@

\begin{table*}
\centering
\label{translationtimes}
\begin{tabular}{|p{0.65\linewidth}|c|} \hline
\textbf{Step} & \textbf{Time (minutes:seconds)} \\ \hline
Front-end                                 
(flow graphs and type inference)          
& 9:01 \\ \hline
LLTyper                                   
(from RPython-level to C-level graphs     
and data)                                 
& 10:38 \\ \hline
Various low-level optimizations           
(convert some heap allocations to local   
variables, inlining, ...)                 
& 6:51 \\ \hline
Database building                         
(this initial back-end step follows all   
graphs and prebuilt data structures       
recursively, assigns names, and orders    
them suitably for code generation)        
& 8:39 \\ \hline
Generating C source                       
& 2:25 \\ \hline
Compiling (\texttt{gcc -O2})              
& 3:23 \\ \hline
\end{tabular}
\caption{dividing overall translation time by stage.}
\end{table*}

@@

>>>>

\begin{figure}
\begin{verbatim}
def ll_dict_insertclean(d, key, value, hash):
  entry = ll_dict_lookup_clean(d, hash)
  ENTRY = lltype.typeOf(entry).TO
  entry.value = value
  entry.key = key
  if hasattr(ENTRY, 'f_hash'): 
    entry.f_hash = hash
  if hasattr(ENTRY, 'f_valid'):    
    entry.f_valid = True
  if hasattr(ENTRY, 'f_everused'): 
    entry.f_everused = True
  d.num_items += 1
  d.num_pristine_entries -= 1
\end{verbatim}
\caption{built-in dict insertion helper.}
\label{fig_insertclean}
\end{figure}

@@

Figure \ref{fig_screenshot}
shows the control flow graph
obtained for a simple function -- this is a screenshot from our graph
viewer, used for debugging; basic block placement is performed by
Graphviz \cite{graphviz}.

@@

However, the bytecode interpreter plays a double role, at two different
levels.  The so-called Object Spaces are \textit{domains} in the abstract
interpretation terminology.  By design, we cleanly separated these
domains from the bytecode interpreter core; the latter is only
responsible for decoding the bytecodes of an application and emulating
the corresponding stack machine.  It treats all actual application-level
objects as black boxes, and dispatches all operations on them to the
Object Space.  The Standard Object Space is a concrete domain, in which
objects are the concrete Python objects of the various built-in types:
lists, dictionaries, and so on.  By opposition, the Flow Object Space is
really an abstract domain.  It handles objects that are placeholders.
Its lattice order (actually a join-semilattice only) is shown in
figure \ref{flowlattice}.

\begin{figure*}
\centering
\includegraphics[scale=0.667]{image/flowlattice.pdf}
\caption{the lattice order of the flow object space.}
\label{flowlattice}
\end{figure*}

This order is extremely simple, because most actual analysis is delayed
to the next phase, the type inference engine.  The objects are either
\textit{Variables}, which are pure placeholders for entirely unknown values,
or \textit{Constants} with a concrete Python object as value.  The order places
Variable as the top, and keeps all \textit{Constants} unordered.  Thus if two
different constants merge during abstract interpretation, we immediately
widen them to Variable.

@@

Figures \ref{latticeoverall} and \ref{latticedetail} show how these
types are ordered to form a lattice -- we generally use its structure of
join-semilattice only, but we will see in section \ref{precision}
a use case for the \textit{meet}.
Not shown on the figures are the list terms, which are simply
unordered with each other, and the Pbcs, which form a
classical finite set-of-subsets lattice.  In addition, we have left
out a number of other annotations that are irrelevant for the basic
description of the annotator and straightforward to handle:
$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
complete list is described in \cite{T}.

\begin{figure}
\centering
\includegraphics[scale=0.667]{image/lattice2.pdf}
\caption{the part about instances and nullable instances, assuming a
  simple class hierarchy with only two direct subclasses of
  \texttt{object}.}
\label{latticedetail}
\end{figure}

@@

\begin{table*}
% the layout of this table isn't wonderful, but it's probably OK.
\centering
\begin{tabular}{|l|c|c|} \hline
\textbf{Interpreter} & 
\textbf{Richards, Time/iteration} &
\textbf{Pystone, Iterations/second} \\ \hline
CPython 2.4.3                &   789ms  (1.0x) & 40322  (1.0x) \\ \hline
pypy-c                       &  4269ms  (5.4x) &  7587  (5.3x) \\ \hline
pypy-c-thread                &  4552ms  (5.8x) &  7122  (5.7x) \\ \hline
pypy-c-stackless             &  5121ms  (6.5x) &  6060  (6.7x) \\ \hline
pypy-c-gcframework           &  6327ms  (8.0x) &  4960  (8.1x) \\ \hline
pypy-c-stackless-gcframework &  8743ms (11.0x) &  3755 (10.7x) \\ \hline
pypy-llvm-c                  &  3797ms  (4.8x) &  7763  (5.2x) \\ \hline
pypy-llvm-c-prof             &  2772ms  (3.5x) & 10245  (3.9x) \\ \hline
\end{tabular}
\caption{summary of interpreter performance.}
\label{perfsumm}
\end{table*}

@@

\begin{table*}
\centering
\begin{tabular}{|p{0.65\linewidth}|c|} \hline
\textbf{Step} & \textbf{Time (minutes:seconds)} \\ \hline
Front-end                                 
(flow graphs and type inference)          
& 9:01 \\ \hline
LLTyper                                   
(from RPython-level to C-level graphs     
and data)                                 
& 10:38 \\ \hline
Various low-level optimizations           
(convert some heap allocations to local   
variables, inlining, ...)                 
& 6:51 \\ \hline
Database building                         
(this initial back-end step follows all   
graphs and prebuilt data structures       
recursively, assigns names, and orders    
them suitably for code generation)        
& 8:39 \\ \hline
Generating C source                       
& 2:25 \\ \hline
Compiling (\texttt{gcc -O2})              
& 3:23 \\ \hline
\end{tabular}
\caption{dividing overall translation time by stage.}
\label{translationtimes}
\end{table*}

@@

\bigskip
